---
title: "Various visualizations"
subtitle: "Supplement for discussion"
author: "Emma Skarstein"
date: "9/6/2020"
output: 
  html_document:
    theme: journal
    code_folding: hide
    number_sections: true
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 12)
```
```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(sf)
library(sp)
library(ggplot2)
library(ggpubr)
library(dplyr)
library(fishualize)
library(hexbin)
library(tidyr)
library(knitr)
library(summarytools)
library(pander)
library(maps)
library(maptools)
library(PointedSDMs)
```

```{r}
trout_artsobs <- readRDS("../R/output/trout_artsobs.RDS")
trout_survey <- readRDS("../R/output/trout_survey.RDS")
trout_artsobs_df <- readRDS("../R/output/trout_artsobs_df.RDS")
trout_survey_df <- readRDS("../R/output/trout_survey_df.RDS")

```
# Looking at the observations

```{r setting up graphic stuff}
# Empty map theme
empty_theme_map <- theme(                              
  plot.background = element_blank(), 
  panel.grid.major = element_blank(), 
  panel.grid.minor = element_blank(), 
  panel.border = element_blank(), 
  panel.background = element_blank(),
  axis.line = element_blank(),
  axis.ticks = element_blank(),
  axis.title.x = element_blank(), 
  axis.title.y = element_blank(),
  axis.text = element_blank()
)
empty_theme <- theme(                              
  plot.background = element_blank(), 
  panel.grid.major = element_blank(), 
  panel.grid.minor = element_blank(), 
  panel.border = element_blank(), 
  panel.background = element_blank(),
  axis.text.x = element_blank(),
  axis.title.x = element_blank()
)

# Setting up color palette
trutta_colors <- fish(n = 20, option = "Salmo_trutta")
#fishualize(option = "Salmo_trutta")

```

## Observation map
```{r}
norway <- ggplot2::map_data("world", region = "Norway(?!:Svalbard)")
norway <- dplyr::setdiff(norway, filter(norway, subregion == "Jan Mayen"))
p1 <- ggplot(data.frame(trout_artsobs)) +
  geom_polygon(data = norway, aes(long, lat, group = group), 
                 color=trutta_colors[11], fill = trutta_colors[11]) +
  geom_point(aes(x = decimalLongitude, y = decimalLatitude), 
               color = trutta_colors[4], alpha = 0.8, size = 1) +
  empty_theme_map +
  ggtitle("Artsobservasjoner")

p2 <- ggplot(data.frame(trout_survey)) +
  geom_polygon(data = norway, aes(long, lat, group = group), 
                 color=trutta_colors[11], fill = trutta_colors[11]) +
  geom_point(aes(x = decimalLongitude, y = decimalLatitude, color = occurrenceStatus), alpha = 0.8, size = 1) +
  scale_color_manual(values=c(trutta_colors[15],trutta_colors[4])) +
  theme(legend.position = "none") +
  empty_theme_map +
  ggtitle("Fish status survey of nordic lakes")
  
ggarrange(p1, p2)
```

Above, I have plotted all the observations for both data sets. Note that the "Fish status survey of nordic lakes" data set is presence/absence, presences are in red in both plots, while absences are in grey when reported.

To get a better impression of observation density across the map, we look at a hexagon map.
```{r}
p1 <- ggplot(data.frame(trout_artsobs), aes(x = decimalLongitude, y = decimalLatitude)) +
  #geom_polygon(data = norway, aes(long, lat, group = group)) +
  geom_hex() +
  empty_theme_map +
  scale_fill_fish(option = "Salmo_trutta", begin = 0.52, end = 0.9, direction = 1)+
  ggtitle("Artsobservasjoner")

p2 <- ggplot(data.frame(trout_survey), aes(x = decimalLongitude, y = decimalLatitude)) +
  #geom_polygon(data = norway, aes(long, lat, group = group)) +
  geom_hex() +
  empty_theme_map +
  scale_fill_fish(option = "Salmo_trutta", begin = 0.52, end = 0.9, direction = 1)+
  ggtitle("Fish status survey of nordic lakes")

ggarrange(p1, p2)

```

## Number of observations per year
```{r obs per year, fig.height=3}
time_counts <- count(trout_artsobs_df, year)

ggplot(time_counts, aes(x = year, y = n)) + 
  geom_line(size = 1, color = trutta_colors[7]) +
  theme_minimal() +
  theme(axis.title = element_blank()) +
  geom_vline(xintercept = 1996, linetype = "dotted", color = trutta_colors[13], size = 1)

```

Number of observations per year for the Artsobservasjoner data. The dotted vertical line is in 1996, which is the year all the survey observations are made.

## Number of observations per square kilometer by county

```{r fylke_barplot}
fylke_areal <- c(4918, 9155, 14912, 48631, 27398, 15438, 14469, 38472, 25192, 454, 4187, 9377, 18623, 15298, 24879, 41898, 7279, 2225)

trout_artsobs_df$county.y <- trout_artsobs_df$county.y %>% recode("Sør-Trøndelag" = "Trøndelag", "Nord-Trøndelag" = "Trøndelag")
dense_fylke <- cbind(count(trout_artsobs_df, vars = county.y), fylke_areal) %>% mutate(obs_per_area = n/fylke_areal)

ggplot(dense_fylke, aes(x = vars, y = obs_per_area)) + 
  geom_bar(stat = "identity", color=trutta_colors[2], fill=trutta_colors[2], width = 0.7) + 
  #geom_text(aes(label = obs_in_county), hjust = -0.5, color="black", size=4) +
  coord_flip() +
 # ylab("Observations per km^2") +
  theme_minimal() + 
  theme(text = element_text(size = 15),
        axis.title.y = element_blank())+
  empty_theme

```
Number of observations per square km, by county, in the Artsobservasjoner data.

# Explanatory variables
The possible explanatory variables are

* `area_km2`: the area of the lake, in square kilometers
* `perimeter_m`: the perimeter of the lake, in meters
* `distance_to_road`: the distance to the closest road
* `eurolst_bio10`: average temperature of the warmest annual quarter, in degrees celcius multiplied by 10
* `catchment_area_km2`: the catchment area of the lake, in square kilometers
* `SCI`: shoreline complexity index
* `HFP`: human footprint index

What is intersting to look at here? Distribution of variables in space, but also distribution of the variables themselves.

```{r}
Covariates <- readRDS("../R/output/Covariates.RDS")
Cov_long <- gather(data.frame(Covariates), key = variable, value = value, area_km2:log_catchment)
```

## Explanatory variables on a map
```{r, fig.height=16}
plot_exp_var <- function(var){
  ggplot(Cov_long %>% dplyr::select(variable, value, decimalLatitude, decimalLongitude) %>% 
                filter(variable==var)) +
    geom_polygon(data = norway, aes(long, lat, group = group), 
                 color = "grey", fill = "grey") +
    geom_point(aes(x = decimalLongitude, y = decimalLatitude, color = value),
             alpha = 0.8, size = 0.3) +
    scale_color_fish(option = "Salmo_trutta") +
    #scale_color_gradient(low = trutta_colors[], high = "darkblue")
    empty_theme_map +
    ggtitle(var)
}
Cov_names <- unique(Cov_long$variable)
var_plots <- lapply(Cov_names, plot_exp_var)

ggarrange(plotlist = var_plots, nrow = 4, ncol = 3)

```

## Explanatory variables summary
```{r covariate table, results = 'asis'}
descr(data.frame(Covariates)[,4:13], stats = "fivenum", transpose = TRUE)
```

## Explanatory variables histograms
To display sort of the same info as the table.

```{r, message=FALSE}
ggplot(Cov_long, aes(x = value)) +
    geom_histogram(fill = trutta_colors[2], color = trutta_colors[2]) +
      facet_wrap(~variable, scales = 'free_x', nrow = 2) +
  theme_minimal() + 
  theme(axis.title = element_blank())
    
```

# Exploring model output
Spatial fields, predictions, comparing preformance of the four models

```{r, warning = FALSE, message=FALSE}
Projection <- CRS("+proj=longlat +ellps=WGS84")
norwayfill <- map("world", "norway", fill=TRUE, plot=FALSE, 
                  ylim=c(58,72), xlim=c(4,32))
IDs <- sapply(strsplit(norwayfill$names, ":"), function(x) x[1])
norway.poly <- map2SpatialPolygons(norwayfill, IDs = IDs, 
                                   proj4string = Projection)
Meshpars <- list(cutoff=0.08, max.edge=c(1, 3), offset=c(1,1))
Mesh <- MakeSpatialRegion(data=NULL, bdry=norway.poly, meshpars=Meshpars,
                          proj = Projection)
source("../R/Model_visualization_functions.R")

```
## Comparing different models

```{r}
res <- readRDS("../R/output/cv_output_4mods.RDS")

dic_values <- matrix(unlist(res), ncol = 5)
av_dic <- rowMeans(dic_values)
```
(please ignore the actual numbers, I do not trust them)

+--------------------+------------------+--------------------+
|                    | 1 spatial field  | 2 spatial fields   |
+====================+==================+====================+
|**Env. covariates** | `r av_dic[1]`           `r av_dic[2]`|
|                    |                                      |
+--------------------+------------------+--------------------+
| **Env. + effort**  | `r av_dic[3]`           `r av_dic[4]`|
| **covariates**     |                                      |
+--------------------+------------------+--------------------+

Above: Result of block cross-validation, comparing four different models. The numbers are the marginal deviance for each model.

## Looking at final model

```{r}
model_final <- readRDS("../R/output/model_final_1.RDS")

summary(model_final$model)
```
We can plot the predictions (not shown) as well as the spatial field means and standard deviations:
```{r}
spat_fields_df <- proj_random_field(model_final$model, sp_polygon = norway.poly, mesh = Mesh$mesh)

ggplot(spat_fields_df) +
  geom_polygon(data = norway, aes(long, lat, group = group), 
               color="white", fill = "white") +
    geom_raster(aes(x = decimalLongitude, y = decimalLatitude, fill = mean)) +
  scale_fill_fish(option = "Salmo_trutta") +
  empty_theme_map +
  facet_wrap(~ field)

ggplot(spat_fields_df) +
  geom_polygon(data = norway, aes(long, lat, group = group), 
               color="white", fill = "white") +
    geom_raster(aes(x = decimalLongitude, y = decimalLatitude, fill = sd)) +
  scale_fill_fish(option = "Salmo_trutta") +
  empty_theme_map +
  facet_wrap(~ field)
```




